---
title: "create figures from simulations"
date: last-modified
---

# accuracy to estimate the GC model parameters

```{r}
#| label: setup
#| echo: false
library(tidyverse)
# setwd("simulations")
columnwidth <- 3 # in inches # 1 point / 72 = 1 inch
```

```{r}
# spec_csv("accuracy_GC_results.csv")
res <- read_csv("accuracy_GC_results.csv", col_types = list(
  topo   = col_character(),
  netT   = col_double(),
  kernel = col_character(),
  lambda = col_double(),
  lmodel = col_character(),
  rep    = col_integer(),
  lambda_hat = col_double(),
  mu_hat = col_double(),
  s2_hat = col_double(),
  v0_hat = col_double(),
  s2_within = col_double(),
  loglik = col_double(),
  aic    = col_double(),
  lrt_pval = col_double(),
  t3_m  = col_double(), t3_v = col_double(),
  t4_m  = col_double(), t4_v = col_double(),
  t6_m  = col_double(), t6_v = col_double(),
  t9_m  = col_double(), t9_v = col_double(),
  t12_m = col_double(),t12_v = col_double(),
  t13_m = col_double(),t13_v = col_double()
))
res <- res |>
  mutate(LRT = factor(
    ifelse(lmodel=="BMw", "BM",
    ifelse(lmodel=="fix1", "fixed λ=1",
    ifelse(lmodel=="true", "true λ",
    ifelse(lrt_pval<0.05, "reject λ=1", "λ=1 fits")))),
    levels=c("reject λ=1","λ=1 fits","fixed λ=1","true λ","BM"))) |>
  mutate(
    lambda_okay = ifelse(lmodel!="estim", TRUE,
      ifelse(lambda==0 & lambda_hat<1e-99, TRUE, # log10(lambda_hat/1e-100) <= 1
      ifelse(abs(log10(lambda_hat/lambda)) <=1, TRUE, FALSE))),
    T = netT,
    netT = factor(netT, levels=c(10,1)),
    lambda_hat_cat = cut(res$lambda_hat,
      breaks=c(-Inf,10^c(-99,-2,-1,-0.5,-0.01,0.01,0.5,1,1.5,2,6))))
```
```{r}
res <- res |>
  mutate(
    eV = # expected Variance within, averaged across the 17 pops
      case_when( # for a tree: use equation for H_u
        topo=="tre" ~ exp(-T) * lambda + (1-exp(-T)),
        # for a network: values calculated in accuracy_GC_simulate.jl
        topo=="net" & T== 1 & lambda== 0 ~ 0.6410849191067046,
        topo=="net" & T== 1 & lambda==.5 ~ 0.8322268597952748,
        topo=="net" & T== 1 & lambda== 1 ~ 1.0233688004838453,
        topo=="net" & T== 1 & lambda== 2 ~ 1.405652681860986,
        topo=="net" & T==10 & lambda== 0 ~ 1.0688460659442562,
        topo=="net" & T==10 & lambda==.5 ~ 1.0691905071381087,
        topo=="net" & T==10 & lambda== 1 ~ 1.0695349483319614,
        topo=="net" & T==10 & lambda== 2 ~ 1.0702238307196663,
      )
    )
```

```{r}
T_labeller = function(x){paste0("T=",x,": ",ifelse(x==1, "high","low")," ILS")}
lmodel_labeller = function(x){
  ifelse(x=="BMw", "BM + σ² within",
  ifelse(x=="fix1", "λ fixed at 1",
  ifelse(x=="estim","λ estimated","λ fixed at true value")))}
kernel_labeller = function(x){
  ifelse(x=="normal", "Brownian motion", "mutational Poisson-Laplace")}
model_labeller = function(x){
  ifelse(x=="BM", "BM + σ² within",
  ifelse(x=="GC1", "GC, λ fixed at 1", "GC, λ estimated"))
}
```

Simulations were done on the 17-taxon network and repeated on its major tree,
with traits simulated either under a mutational Poisson-Laplace model
or a Brownian motion (along each gene tree). The results are very similar,
so we restrict the results to the network (with reticulation) and mutational
model only, to present in the manuscript, for simplicity.

```{r}
#| label: "filter results to network & mutational model"
res_nm <- filter(res, topo=="net" & kernel=="mutational")
```

evolutionary rate σ² is well estimated under low ILS and GC model:

```{r}
p <- ggplot(res_nm,
  aes(y=s2_hat, x=as.factor(lambda))) +
  geom_hline(yintercept=1) + # 1 = true σ²
  geom_boxplot(outliers=FALSE) +
  geom_jitter(aes(color=lambda_hat_cat), width=.3, size=.1, stroke = 0) +
  # scale_color_grey(start=0.6, end=0) +
  scale_color_brewer(palette="Spectral", direction=-1) +
  facet_grid(netT ~ lmodel, scales = "free_y",
    labeller = labeller(netT=T_labeller, lmodel=lmodel_labeller)) +
  xlab("true λ") + ylab("estimated σ²") +
  scale_y_sqrt() +
  theme_bw() +
  theme(legend.position="none",
        text=element_text(size=unit(9, 'pt')),)
p
```

```{r}
ggsave("accuracy_GC_sigma2.pdf", p,
  device = cairo_pdf, 
  width = 2*columnwidth, height = 1.2*columnwidth, unit = "in")
```


```{r}
p <- ggplot(res_nm,
  aes(y=mu_hat, x=as.factor(lambda))) +
  geom_hline(yintercept=0) + # 0 = true population mean at the root
  geom_boxplot(outliers=FALSE) +
  geom_jitter(color=grey(.8), width=.3, size=.1, stroke=0) +
  scale_y_continuous(minor_breaks=seq(-5,5,by=1)) +
  facet_grid(netT ~ lmodel,
    labeller = labeller(netT=T_labeller, lmodel=lmodel_labeller)) +
  xlab("true λ") + ylab("estimated m₀") +
  theme_bw() +
  theme(legend.position="none", text=element_text(size=unit(9,'pt')))
p
```

```{r}
ggsave("accuracy_GC_mu.pdf", p,
    device = cairo_pdf,
    width = 2*columnwidth, height = 1.2*columnwidth, unit="in")
```


The estimation of λ is quite poor:

using data from both kernels (mutational and BM), we see a
very slightly better estimation of λ from BM traits:

```{r}
p <- ggplot(
    filter(res, topo=="net" & lmodel=="estim"),
    aes(y=lambda_hat, x=as.factor(lambda))) +
  geom_boxplot(outliers=FALSE) +
  scale_y_continuous(transform="log1p", minor_breaks=NULL,
    breaks=c(0,0.1,0.5,1,2,5,10,100,200,400,600)) +
  geom_jitter(aes(color=LRT), width=0.1, size=0.2) +
  scale_color_manual(
    values = c("black", "sienna1"),
    guide = guide_legend(reverse=TRUE,
        title="likelihood ratio test",
        override.aes = list(size=1))) +
  facet_grid(netT ~ kernel, labeller = labeller(
    kernel=kernel_labeller, netT=T_labeller)) +
  xlab("true λ") + ylab("estimated λ") +
  theme_bw() +
  theme(legend.position="bottom")
p
```

```{r}
ggsave("accuracy_GC_lambda_2kernels.pdf", p,
    device = cairo_pdf, width=6, height=6)
```

but the effect is very slight, and not worth using space
in the main part of the paper.
below: figure for the mutational model only.

```{r}
lbreaks = c(0,0.5,1,2,5,10,100,200,400,600)
p <- ggplot(filter(res_nm, lmodel=="estim"),
    aes(y=lambda_hat, x=as.factor(lambda))) +
  geom_boxplot(outliers=FALSE) +
  scale_y_continuous(transform="log1p", minor_breaks=NULL,
    breaks=lbreaks, labels=as.character(lbreaks)) +
  geom_jitter(aes(color=LRT), width=0.1, size=0.2) +
  scale_color_manual(
    values = c("black", "sienna1"),
    guide = guide_legend(reverse=TRUE,
        title="likelihood ratio test",
        override.aes = list(size=1))) +
  facet_grid(~ netT, labeller = labeller(netT=T_labeller)) +
  xlab("true λ") + ylab("estimated λ") +
  theme_bw() +
  theme(legend.position="bottom", text=element_text(size=unit(9,'pt')))
p
```

```{r}
ggsave("accuracy_GC_lambda.pdf", p,
    device = cairo_pdf, 
    width = 1.5*columnwidth, height = 0.9*columnwidth, unit="in")
```

Model selection with AIC:
including the BM with intra-specific variation, but
excluding the GC model with the true λ because
we never know the true λ in practice (and when λ=1,
the CG model with the true λ and λ fixed to 1 are the same).

```{r}
aic_df <- res |>
  select(topo, netT, kernel, lambda, rep, lmodel, aic) |>
  pivot_wider(names_from="lmodel", values_from="aic") |>
  mutate(
  bestmodel = case_when(
      fix1 <= estim & fix1 <= BMw ~ "fix1",
      BMw <= estim ~ "BMw",
      TRUE ~ "estim" )) |> mutate(
  min_aic = case_when(
      bestmodel == "fix1" ~ fix1,
      bestmodel == "estim" ~ estim,
      bestmodel == "BMw" ~ BMw )) |> mutate(
  delta_fix1  = fix1  - min_aic,
  delta_estim = estim - min_aic,
  delta_BMw   = BMw   - min_aic)
mean(aic_df$delta_fix1 == 0)     # 0.7809375
mean(aic_df$bestmodel == "fix1") # same
# mean(aic_df$bestmodel == "true") # 0.57 when included
mean(aic_df$bestmodel == "BMw") # 0.1809375
# 0.14625 when 'true' was included
mean(aic_df$BMw < aic_df$fix1) # 0.1859375
```

```{r}
aic_res <- aic_df |> filter(topo=="net" & kernel=="mutational") |>
  group_by(netT,lambda) |>
  summarize(
    n_tot = n(),
    n_GC1 = sum(bestmodel=="fix1"),
    n_BM = sum(bestmodel=="BMw"),
    n_GCestim = sum(bestmodel=="estim"),
    .groups = "drop"
  )
aic_res_long <- aic_res |> select(!n_tot) |>
  pivot_longer(!c("netT","lambda"), values_to="n",
  names_to="model", names_prefix="n_") |>
  mutate(model = factor(model, levels=c("BM","GCestim","GC1")))
aic_res
```

  netT  lambda n_tot n_GC1  n_BM n_GCestim
  10       0     100    82    17         1
  10       0.5   100    81    16         3
  10       1     100    85     7         8
  10       2     100    76    19         5
  1        0     100    73    26         1
  1        0.5   100    69    28         3
  1        1     100    85    12         3
  1        2     100    81    13         6

```{r}
p <- ggplot(aic_res_long,
  aes(x=as.factor(lambda), y=n, fill=model)) +
  geom_bar(stat="identity") +
  scale_y_continuous(limits=c(0,100), expand=0) +
  scale_fill_manual(
    values = c("sienna4", "burlywood3", "burlywood1"),
    name = "model favored by AIC:",
    labels=c("BM + σ² within", "GC, λ estimated", "GC, λ fixed at 1")) +
  facet_grid(~ netT, labeller = labeller(netT=T_labeller)) +
  xlab("true λ") +
  ylab("frequency favored by AIC") +
  theme_bw() +
  theme(text=element_text(size=unit(9, 'pt')))
p
```

```{r}
ggsave("accuracy_GC_aic.pdf", p,
      device = cairo_pdf, 
      width = 1.6*columnwidth, height = 0.8*columnwidth, unit = "in")
```

λ=1 is almost never rejected:
```{r}
res |> filter(lmodel=="estim") |>
  group_by(lambda, netT) |>
  summarize(
    nreject=sum(LRT=="reject λ=1"),
    naccept=sum(LRT=="λ=1 fits"),
    ntotal=n()) |>
  mutate(power = nreject/ntotal) |> print(n=32)
```

    λ   T    nreject naccept ntotal  power
    0   10         2     398    400 0.005 
    0   1          0     400    400 0     
    0.5 10         2     398    400 0.005 
    0.5 1          3     397    400 0.0075
    1   10        13     387    400 0.0325
    1   1          3     397    400 0.0075
    2   10         8     392    400 0.02  
    2   1          9     391    400 0.0225

restricted to the network, mutational kernel:
```{r}
res_nm |> filter(lmodel=="estim") |>
  # group_by(lambda, netT, topo, kernel) |>
  group_by(netT) |>
  summarize(
    nreject=sum(LRT=="reject λ=1"),
    naccept=sum(LRT=="λ=1 fits"),
    ntotal=n()) |>
  mutate(power = nreject/ntotal) |> print(n=32)
```

    netT  nreject naccept ntotal  power
    10          8     392    400 0.02  
    1           1     399    400 0.0025

"λ=1" is rejected only 9/800 = 1.125% of the time overall

correlation between λ and σ² when λ is estimated:
I would have expected the estimates of
v0 and σ² to be correlated, but they are not particularly when T=10:

```{r}
p <- ggplot(filter(res_nm, lmodel=="estim"),
    aes(y=lambda_hat, x=s2_hat, color=as.factor(lambda))) +
  scale_y_continuous(transform="log1p", minor_breaks=NULL,
    breaks=c(0,0.5,1,2,5,10,100,200,400,600)) +
  geom_point(size=0.2) +
  scale_color_manual(
    values = c("grey", "black", "sienna1", "sienna4"),
    guide = guide_legend(title="true λ",
        override.aes = list(size=1))) +
  facet_grid( ~ netT,
    labeller = labeller(netT=T_labeller)) +
  xlab("estimated σ²") + ylab("estimated λ") +
  theme_bw() +
  theme(legend.position="bottom")
p
```
```{r}
ggsave("accuracy_GC_lambda_sigma2.pdf", p,
    device = cairo_pdf, width=6, height=4)
```


Sanity checks: under GC `v0_hat = lambda_hat * s2_hat` and
if the phylogeny is a tree then `s2_within` matches the formula for Hu.

```{r}
#| eval: false
subset(res, lmodel == "estim") |> # aligned: good
  ggplot(aes(y=v0_hat, x = lambda_hat * s2_hat)) + geom_point()
subset(res, lmodel!="BMw" & topo=="tre") |> # aligned: good
  ggplot(aes(y=s2_within, x=exp(-T)*v0_hat + (1-exp(-T))*s2_hat)) + geom_point()
```


v0 = λσ² is estimated better under high ILS than low ILS:
```{r}
p <- ggplot(subset(res_nm, lmodel!="BMw"),
    aes(y=v0_hat, x=as.factor(lambda))) +
  geom_boxplot(outliers=FALSE) +
  geom_jitter(aes(color=lambda_hat_cat), width=.3, size=.1, stroke = 0) +
  scale_y_continuous(
    transform="log1p", minor_breaks=NULL,
    breaks=c(0,0.1,0.5,1,2,5,10,100,200,400,600)) +
  scale_color_brewer(palette="Spectral", direction=-1,
    guide = NULL,
    ) +
  facet_grid(netT ~ lmodel, scales = "free_y",
    labeller = labeller(netT=T_labeller, lmodel=lmodel_labeller)) +
  xlab("true λ") + ylab("estimated v0 = σ²λ (true=λ)") +
  theme_bw() +
  theme(legend.position="bottom",
    legend.margin=margin(-10,0,0,0),
    text=element_text(size=unit(9, 'pt')))
p
```

```{r}
ggsave("accuracy_GC_v0.pdf", p,
    device = cairo_pdf,
    width = 2*columnwidth, height = 1.2*columnwidth, unit = "in")
```


The within-species variance (a weighted average of v0 and σ² under a tree)
is close to v0 under high ILS and to σ² under low ILS, roughly.

It is well estimated under low ILS:

```{r}
eVs <- res |> group_by(topo, netT, lambda) |>
  summarize(eV = first(eV), .groups="drop")
```

```{r}
p <- ggplot(subset(res, netT=="10"),
  aes(y=s2_within/eV, x=as.factor(lambda))) +
  geom_boxplot(outliers=FALSE) +
  geom_jitter(aes(color=lambda_hat_cat), width=.3, size=.2) +
  scale_color_brewer(palette="Spectral", direction=-1,
    guide = NULL,
    ) +
  facet_grid(kernel ~ lmodel,
    labeller = labeller(kernel=kernel_labeller, lmodel=lmodel_labeller)) +
  xlab("true λ") + ylab("ratio: fitted/true expected within-population variance") +
  theme_bw() +
  theme(legend.position="bottom",
    legend.margin=margin(-10,0,0,0))
p
```
```{r}
ggsave("accuracy_GC_eV_2kernels_T10.pdf", p,
    device = cairo_pdf, width=8, height=6)
```

It is also well estimated under high ILS regardless of the method
(e.g. even if λ is estimated), and without visible bias:

```{r}
p <- ggplot(subset(res, netT=="1"),
  aes(y=s2_within/eV, x=as.factor(lambda))) +
  geom_boxplot(outliers=FALSE) +
  geom_jitter(aes(color=lambda_hat_cat), width=.3, size=.2) +
  scale_color_brewer(palette="Spectral", direction=-1,
    guide = NULL,
    ) +
  facet_grid(kernel ~ lmodel,
    labeller = labeller(kernel=kernel_labeller, lmodel=lmodel_labeller)) +
  xlab("true λ") + ylab("ratio: fitted/true expected within-population variance") +
  theme_bw() +
  theme(legend.position="bottom",
    legend.margin=margin(-10,0,0,0))
p
```
```{r}
ggsave("accuracy_GC_eV_2kernels_T1.pdf", p,
    device = cairo_pdf, width=8, height=6)
```

for the paper, focus on the network with mutational model,
showing both low & high ILS:

```{r}
p <- ggplot(res_nm,
  aes(y=s2_within/eV, x=as.factor(lambda))) +
  geom_boxplot(outliers=FALSE) +
  geom_jitter(aes(color=lambda_hat_cat), width=.3, size=.1, stroke = 0) +
  scale_color_brewer(palette="Spectral", direction=-1,
    guide = NULL) +
  facet_grid(netT ~ lmodel,
    labeller = labeller(netT=T_labeller, lmodel=lmodel_labeller)) +
  xlab("true λ") + ylab("ratio: fitted/true expected within-population variance") +
  theme_bw() +
  theme(legend.position="bottom",
    legend.margin=margin(-10,0,0,0),
    text=element_text(size=unit(9, 'pt')))
p
```
```{r}
ggsave("accuracy_GC_eV.pdf", p,
    device = cairo_pdf, 
    width = 2*columnwidth, height = 1.2*columnwidth, unit = "in")
```

```{r}
res_nm |> mutate(s2_within_ratio = s2_within/eV) |>
  summarize(m = mean(s2_within_ratio), s = sd(s2_within_ratio))
```

unbiased: m = 1.001922, s=0.0917733
SE: 0.0917733/sqrt(3200) = 0.001622338

Out of curiosity: look at the simulated mean & variance
within some of the populations.
t3 and t4 are not below *any* reticulation.
t6, t9, t12, t13 are each below 1 reticulation,
the most recent reticulation being that above t9.

```{r}
res |> filter(!is.na(t3_m)) |>
  group_by(topo, netT, lambda) |>
  summarize(
    eV = first(eV),
    t3_mV = mean(t3_v),
    t4_mV = mean(t4_v),
    t6_mV = mean(t6_v),
    t9_mV = mean(t9_v),
    t12_mV = mean(t12_v),
    t13_mV = mean(t13_v),
    #t3_vM = var( t3_m),
    #t4_vM = var( t4_m),
    #t6_vM = var( t6_m),
    #t9_vM = var( t9_m),
    )
```

    topo   T  λ     eV  t3_mV t4_mV t6_mV t9_mV t12_mV t13_mV
    net   10  0   1.07  1.04  1.03  0.995 2.07  0.994  0.984
    net   10  0.5 1.07  0.986 0.978 0.967 2.14  0.993  1.06 
    net   10  1   1.07  1.02  1.01  1.01  2.19  1.03   1.01 
    net   10  2   1.07  0.999 1.02  1.03  2.10  1.01   1.02 
    net   1   0   0.641 0.627 0.642 0.653 0.716 0.657  0.629
    net   1   0.5 0.832 0.812 0.834 0.857 0.959 0.859  0.833
    net   1   1   1.02  1.03  0.991 1.06  1.25  1.06   1.09 
    net   1   2   1.41  1.36  1.38  1.42  1.70  1.45   1.54 
    tre   10  0   1.000 0.996 0.997 1.00  0.972 0.998  0.986
    tre   10  0.5 1.000 1.01  1.04  1.00  0.976 1.01   0.963
    tre   10  1   1     1.00  1.03  0.982 0.978 1.01   0.956
    tre   10  2   1.00  0.968 1.05  0.946 0.988 1.00   1.01 
    tre   1   0   0.632 0.651 0.625 0.640 0.639 0.638  0.639
    tre   1   0.5 0.816 0.814 0.826 0.865 0.815 0.804  0.832
    tre   1   1   1     1.000 1.01  0.957 1.00  0.997  0.996
    tre   1   2   1.37  1.34  1.33  1.37  1.41  1.33   1.33
