---
title: "create figures from simulations on the tomato tree"
date: last-modified
---

# accuracy of Gaussian with ILS: tomato tree

many loci: so the joint Gaussian assumption should be a good approximation.

```{r}
#| label: setup
#| echo: false
library(tidyverse)
setwd("simulations")
columnwidth <- 3 # in inches # 1 point / 72 = 1 inch
```

```{r}
# spec_csv("accuracy_GC_results.csv")
res <- read_csv("accuracy_GC_results_tomato.csv", col_types = list(
  topo   = col_character(),
  netT   = col_double(),
  kernel = col_character(),
  lambda = col_double(),
  lmodel = col_character(),
  rep    = col_integer(),
  lambda_hat = col_double(),
  mu_hat = col_double(),
  s2_hat = col_double(),
  v0_hat = col_double(),
  s2_within = col_double(),
  loglik = col_double(),
  aic    = col_double(),
  lrt_pval = col_double(), # default has trouble with this!
  t3_m  = col_double(), t3_v = col_double(),
  t4_m  = col_double(), t4_v = col_double(),
  t6_m  = col_double(), t6_v = col_double(),
  t9_m  = col_double(), t9_v = col_double(),
  t12_m = col_double(),t12_v = col_double(),
  t13_m = col_double(),t13_v = col_double()
))
res <- res |>
  mutate(LRT = factor(
    ifelse(lmodel=="BMw", "BM",
    ifelse(lmodel=="fix1", "fixed λ=1",
    ifelse(lmodel=="true", "true λ",
    ifelse(lrt_pval<0.05, "reject λ=1", "λ=1 fits")))),
    levels=c("reject λ=1","λ=1 fits","fixed λ=1","true λ","BM"))) |>
  mutate(
    lambda_okay = ifelse(lmodel!="estim", TRUE,
      ifelse(lambda==0 & lambda_hat<1e-99, TRUE, # log10(lambda_hat/1e-100) <= 1
      ifelse(abs(log10(lambda_hat/lambda)) <=1, TRUE, FALSE))),
    T = netT,
    netT = factor(netT, levels=c(6,1)),
    lambda_hat_cat = cut(res$lambda_hat,
      breaks=c(-Inf,10^c(-99,-2,-1,-0.5,-0.01,0.01,0.5,1,1.5,2,6))))
```
```{r}
res <- res |>
  mutate(
    eV = # expected Variance within, averaged across the 17 pops
      case_when( # for a tree: use equation for H_u
        topo=="tre" ~ exp(-T) * lambda + (1-exp(-T)),
      )
    )
```

```{r}
T_labeller = function(x){paste0("T=",x,": ",ifelse(x==1, "high","low")," ILS")}
param_labeller = function(x){ifelse(x=="sd_hat", "estimated σ", "estimated m₀")}
lmodel_labeller = function(x){
  ifelse(x=="BMw", "BM + σ² within",
  ifelse(x=="fix1", "λ fixed at 1",
  ifelse(x=="estim","λ estimated","λ fixed at true value")))}
kernel_labeller = function(x){
  ifelse(x=="normal", "Brownian motion", "mutational Poisson-Laplace")}
model_labeller = function(x){
  ifelse(x=="BM", "BM + σ² within",
  ifelse(x=="GC1", "GC, λ fixed at 1", "GC, λ estimated"))
}
```

Simulations were done on the 12-taxon tomato tree,
with traits simulated either a mutational Poisson-Laplace model. 

```{r}
#| label: "filter results to network & mutational model"
res_tm <- filter(res, topo=="tre" & kernel=="mutational")
```

evolutionary rate σ² is well estimated under low ILS (empirical tree) and GC model:

```{r}
p <- ggplot(res_tm,
  aes(y=s2_hat, x=as.factor(lambda))) +
  geom_hline(yintercept=1) + # 1 = true σ²
  geom_boxplot(outliers=FALSE) +
  geom_jitter(aes(color=lambda_hat_cat), width=.3, size=.1, stroke = 0) +
  # scale_color_grey(start=0.6, end=0) +
  scale_color_brewer(palette="Spectral", direction=-1) +
  facet_grid(netT ~ lmodel, scales = "free_y",
    labeller = labeller(netT=T_labeller, lmodel=lmodel_labeller)) +
  xlab("true λ") + ylab("estimated σ²") +
  scale_y_sqrt() +
  theme_bw() +
  theme(legend.position="none",
        text=element_text(size=unit(9, 'pt')),)
p
```

```{r}
ggsave("accuracy_GC_sigma2_tomato_1_6.pdf", p,
  device = cairo_pdf, 
  width = 2*columnwidth, height = 1.2*columnwidth, unit = "in")
```

plot with empirical tree only (total heigh 6 coal. units):
```{r}
p <- ggplot(subset(res_tm, netT == 6),
  aes(y=s2_hat, x=as.factor(lambda))) +
  geom_hline(yintercept=1) + # 1 = true σ²
  geom_boxplot(outliers=FALSE) +
  geom_jitter(aes(color=lambda_hat_cat), width=.3, size=.1, stroke = 0) +
  # scale_color_grey(start=0.6, end=0) +
  scale_color_brewer(palette="Spectral", direction=-1) +
  facet_grid(. ~ lmodel, scales = "free_y",
    labeller = labeller(netT=T_labeller, lmodel=lmodel_labeller)) +
  xlab("true λ") + ylab("estimated σ²") +
  scale_y_sqrt() +
  theme_bw() +
  theme(legend.position="none",
        text=element_text(size=unit(9, 'pt')),)
p
```

```{r}
ggsave("accuracy_GC_sigma2_tomato.pdf", p,
  device = cairo_pdf, 
  width = 2*columnwidth, height = 0.7*columnwidth, unit = "in")
```

numerical values for the bias, precision and accuracy in estimating σ
(note: square-root of σ², true is 1):

```{r}
s2_hat_summary <- res_tm |>
  mutate(s_hat = sqrt(s2_hat)) |>
  group_by(netT, lmodel, lambda) |>
  summarize(
    bias = mean(s_hat)-1,
    stde = sqrt(mean((s_hat - mean(s_hat))^2)),
    rmse = sqrt(mean((s_hat-1)^2)),
    .groups = "drop_last"
  )
# with(s2_hat_summary, max((bias^2 + stde^2)/rmse^2 -1)) # 4.440892e-16
write_csv(s2_hat_summary, "accuracy_GC_results_s2summary_tomato.csv")
s2_hat_summary |> summarize(
  bias = mean(abs(bias)),
  stde = mean(stde),
  rmse = mean(rmse),
  .groups = "drop"
)
```

  netT  lmodel    bias  stde  rmse
1 6     BMw    0.0456  0.265 0.269
2 6     estim  0.0284  0.124 0.127
3 6     fix1   0.0164  0.118 0.120
4 6     true   0.0107  0.118 0.119
5 1     BMw    0.237   0.369 0.461
6 1     estim  0.114   0.432 0.449
7 1     fix1   0.131   0.121 0.192
8 1     true   0.00657 0.124 0.124

Ancestral mean is well estimated:

```{r}
p <- ggplot(res_tm,
  aes(y=mu_hat, x=as.factor(lambda))) +
  geom_hline(yintercept=0) + # 0 = true population mean at the root
  geom_boxplot(outliers=FALSE) +
  geom_jitter(color=grey(.8), width=.3, size=.1, stroke=0) + # width=0.1, size=0.2
  scale_y_continuous(minor_breaks=seq(-5,5,by=1)) +
  facet_grid(netT ~ lmodel,
    labeller = labeller(netT=T_labeller, lmodel=lmodel_labeller)) +
  xlab("true λ") + ylab("estimated m₀") +
  theme_bw() +
  theme(legend.position="none", text=element_text(size=unit(9,'pt')))
p
```

```{r}
ggsave("accuracy_GC_mu_tomato.pdf", p,
    device = cairo_pdf,
    width = 2*columnwidth, height = 1.2*columnwidth, unit="in")
```


The estimation of λ is quite poor:

```{r}
lbreaks = c(0,0.5,1,2,5,10,100,200,400,600)
p <- ggplot(filter(res_tm, lmodel=="estim" & netT==6),
    aes(y=lambda_hat, x=as.factor(lambda))) +
  geom_boxplot(outliers=FALSE) +
  scale_y_continuous(transform="log1p", minor_breaks=NULL,
    breaks=lbreaks, labels=as.character(lbreaks)) +
  geom_jitter(aes(color=LRT), width=0.1, size=0.2) +
  scale_color_manual(
    values = c("black", "sienna1"),
    guide = guide_legend(reverse=TRUE,
        title="likelihood ratio test",
        override.aes = list(size=1))) +
  facet_grid(~ netT, labeller = labeller(netT=T_labeller)) +
  xlab("true λ") + ylab("estimated λ") +
  theme_bw() +
  theme(text=element_text(size=unit(9,'pt')))
p
```

```{r}
ggsave("accuracy_GC_lambda_tomato.pdf", p,
    device = cairo_pdf, 
    width = 1.2*columnwidth, height = 0.9*columnwidth, unit="in")
```

```{r}
res_tm |> filter(lmodel=="estim") |>
  # group_by(lambda, netT, topo, kernel) |>
  group_by(netT) |>
  summarize(
    nreject=sum(LRT=="reject λ=1"),
    naccept=sum(LRT=="λ=1 fits"),
    ntotal=n()) |>
  mutate(power = nreject/ntotal) |> print(n=32)
```

    netT  nreject naccept ntotal  power
1 6           3     397    400 0.0075
2 1           1     399    400 0.0025

"λ=1" is rejected only 3/400 = 0.75% of the time overall

Plot mu and sigma2 parameters in a single graph

```{r}
res_tm1 <- subset(res_tm, netT == 6)
res_tm1$sd_hat <- sqrt(res_tm1$s2_hat)
res_tm1_long <- pivot_longer(res_tm1, cols = c("mu_hat", "sd_hat"), names_to = "parameter")
res_tm1_long$true_value <- 1
res_tm1_long[res_tm1_long$parameter == "mu_hat",]$true_value <- 0
res_tm1_long$parameter <- factor(res_tm1_long$parameter, levels = c("sd_hat", "mu_hat"))
```
```{r}
p <- ggplot(res_tm1_long,
  aes(y=value, x=as.factor(lambda))) +
  geom_hline(aes(yintercept=true_value)) + # 1 = true σ²
  geom_boxplot(outliers=FALSE) +
  geom_jitter(aes(color=lambda_hat_cat), width=.3, size=.1, stroke = 0) +
  # scale_color_grey(start=0.6, end=0) +
  scale_color_brewer(palette="Spectral", direction=-1) +
  facet_grid(parameter ~ lmodel, scales = "free_y", switch = "y",
    labeller = labeller(parameter=param_labeller, lmodel=lmodel_labeller)) +
  xlab("true λ") +
  theme_bw() +
  theme(legend.position="none",
        axis.title.y = element_blank(),
        text=element_text(size=unit(9, 'pt')),
        title = element_text(size = 7),
        strip.text.y = element_text(vjust = 0),
        strip.placement = "outside",
        strip.background = element_blank(),
  )
p
```

```{r}
ggsave("accuracy_GC_sigma2_mu_tomato.pdf", p,
  device = cairo_pdf, 
  width = 2*columnwidth, height = 1.2*columnwidth, unit = "in")
```

Model selection with AIC:
including the BM with intra-specific variation, but
excluding the GC model with the true λ because
we never know the true λ in practice (and when λ=1,
the CG model with the true λ and λ fixed to 1 are the same).

```{r}
aic_df <- res_tm1 |>
  select(topo, netT, kernel, lambda, rep, lmodel, aic) |>
  pivot_wider(names_from="lmodel", values_from="aic") |>
  mutate(
  bestmodel = case_when(
      fix1 <= estim & fix1 <= BMw ~ "fix1",
      BMw <= estim ~ "BMw",
      TRUE ~ "estim" )) |> mutate(
  min_aic = case_when(
      bestmodel == "fix1" ~ fix1,
      bestmodel == "estim" ~ estim,
      bestmodel == "BMw" ~ BMw )) |> mutate(
  delta_fix1  = fix1  - min_aic,
  delta_estim = estim - min_aic,
  delta_BMw   = BMw   - min_aic)
mean(aic_df$delta_fix1 == 0)     # 0.7809375
mean(aic_df$bestmodel == "fix1") # same
# mean(aic_df$bestmodel == "true") # 0.57 when included
mean(aic_df$bestmodel == "BMw") # 0.1809375
# 0.14625 when 'true' was included
mean(aic_df$BMw < aic_df$fix1) # 0.1859375
mean(aic_df$BMw < aic_df$estim) # 0.4846875
```

```{r}
aic_res <- aic_df |> filter(topo=="tre" & kernel=="mutational") |>
  group_by(netT,lambda) |>
  summarize(
    n_tot = n(),
    n_GC1 = sum(bestmodel=="fix1"),
    n_BM = sum(bestmodel=="BMw"),
    n_GCestim = sum(bestmodel=="estim"),
    .groups = "drop"
  )
aic_res_long <- aic_res |> select(!n_tot) |>
  pivot_longer(!c("netT","lambda"), values_to="n",
  names_to="model", names_prefix="n_") |>
  mutate(model = factor(model, levels=c("BM","GCestim","GC1")))
aic_res
```

  netT  lambda n_tot n_GC1  n_BM n_GCestim
1 6        0     100    78    19         3
2 6        0.5   100    81    15         4
3 6        1     100    81    16         3
4 6        2     100    82    11         7
```{r}
colSums(aic_res[, 3:6]) / 400
```

    n_tot     n_GC1      n_BM n_GCestim 
   1.0000    0.8050    0.1525    0.0425
   
```{r}
p <- ggplot(aic_res_long,
  aes(x=as.factor(lambda), y=n, fill=model)) +
  geom_bar(stat="identity") +
  scale_y_continuous(limits=c(0,100), expand=0) +
  scale_fill_manual(
    values = c("sienna4", "burlywood3", "burlywood1"),
    name = "model favored by AIC:",
    labels=c("BM + σ² within", "GC, λ estimated", "GC, λ fixed at 1")) +
  facet_grid(~ netT, labeller = labeller(netT=T_labeller)) +
  xlab("true λ") +
  ylab("frequency favored by AIC") +
  theme_bw() +
  theme(#legend.position="bottom",
        text=element_text(size=unit(9, 'pt')))
p
```

```{r}
ggsave("accuracy_GC_aic_tomato.pdf", p,
      device = cairo_pdf, 
      width = 1.2*columnwidth, height = 0.8*columnwidth, unit = "in")
```

Best model when comparing only BMw and lambda esitm
```{r}
aic_res_BMGC <- aic_df |> filter(topo=="tre" & kernel=="mutational") |>
  group_by(netT) |>
  summarize(
    n_tot = n(),
    n_BM = mean(BMw<estim),
    n_GCestim = mean(BMw>=estim),
    .groups = "drop"
  )
aic_res_BMGC
```

  netT  n_tot  n_BM n_GCestim
1 6       400 0.615     0.385
2 1       400 0.645     0.355

λ=1 is almost never rejected:
```{r}
res |> filter(lmodel=="estim") |>
  # group_by(lambda, netT, topo, kernel) |>
  group_by(lambda, netT) |>
  summarize(
    nreject=sum(LRT=="reject λ=1"),
    naccept=sum(LRT=="λ=1 fits"),
    ntotal=n()) |>
  mutate(power = nreject/ntotal) |> print(n=32)
```

  lambda netT  nreject naccept ntotal power
1    0   6           0     100    100  0   
2    0   1           0     100    100  0   
3    0.5 6           1      99    100  0.01
4    0.5 1           0     100    100  0   
5    1   6           0     100    100  0   
6    1   1           0     100    100  0   
7    2   6           2      98    100  0.02
8    2   1           1      99    100  0.01
